---
title: "Reproducible Research- Peer Assessment 1"
author: "Bozidarka Jankovic"
date: "July 19, 2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment="")
```

## Introduction

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data
The first step is to load the data and required packages. It is important to be sure that you have downloaded the activity file from the course site. Also we will download dplyr library, as it is necessary for the work.
```{r}
data_row<-read.csv("activity.csv")
```

```{r}
library(dplyr)
```

##What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
# clean row data set of NA values
data<-na.omit(data_row)
head(data, 20)
```

```{r}
#group data set per day
data_group<-group_by(data,date)
total_number_steps_per_date<-summarise(data_group,total_steps_per_day=sum(steps))
total_number_steps_per_date
```

Create a histogram of the total number of steps taken each day

```{r}
hist(total_number_steps_per_date$total_steps_per_day,main="Histogram of total number of steps per day", 
     xlab="Total number of steps in a day")
```

Calculate and report the mean and median total number of steps taken per day

```{r}
# mean of the total number of steps taken per day
mean_steps_per_day<-mean(total_number_steps_per_date$total_steps_per_day)
mean_steps_per_day
```
```{r}
# mediana of the total number of steps taken per day
mediana_steps_per_day<-median(total_number_steps_per_date$total_steps_per_day)
mediana_steps_per_day
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
#calulate average number of steps taken per inteval
#group data per inteval
data_group_interval<-group_by(data,interval)
#summarise avearge numebr of steps per interval
intervalav_steps<-summarise(data_group_interval,avr_steps_interval=mean(steps))
intervalav_steps
```

```{r}
plot(intervalav_steps$interval ,intervalav_steps$avr_steps_interval, type="l",
     xlab="Interval",
     ylab="Average steps taken",
     main="Average steps taken during 5 minute interval")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
intervalav_steps$interval[which.max(intervalav_steps$avr_steps_interval)]
```

##Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
#Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
total_numer_NA<-sum(is.na(data_row) == TRUE)
total_numer_NA
```

```{r}
#Strategy for filling in all of the missing values in the dataset. 
filldata<-data_row
for (i in 1:nrow(filldata)) {
  if(is.na(filldata$steps[i]) == TRUE)
  {
    tmp<-intervalav_steps$avr_steps_interval[which(intervalav_steps$interval==filldata$interval[i])]
    filldata$steps[i]<-tmp
    
  }
}
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
filled_missing_data<-group_by(filldata,date)

total_number_steps_per_date_after_cleaning<-summarise(filled_missing_data,total=sum(steps))
total_number_steps_per_date_after_cleaning
```

```{r}
hist(total_number_steps_per_date_after_cleaning$total,main="Histogram of total number of steps per day after filling missing values", 
     xlab="Total number of steps in a day")
```

```{r}
mean(total_number_steps_per_date_after_cleaning$total)
median(total_number_steps_per_date_after_cleaning$total)
```

Conclusion:Comparing original and modified data There is no significant change in the results: the number of valid samples in the modified data is greater, so the total steps is also greater, but the histogram is very similar in shape, and the mean and median are practily the same

## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
week_day <- function(date) {
  wd <- weekdays(as.Date(date, '%Y-%m-%d'))
  if  (!(wd == 'Saturday' || wd == 'Sunday')) {
    x <- 'Weekday'
  } else {
    x <- 'Weekend'
  }
  x
}
filled_missing_data$type_of_day<-unlist(lapply(filled_missing_data$date,week_day))
```

```{r}
data_group_interval<-group_by(filled_missing_data,type_of_day,interval)
data_group_interval
intervalav_steps<-summarise(data_group_interval,avr_steps_interval=mean(steps))
intervalav_steps
```

```{r}
library(ggplot2)

qplot(interval, 
      avr_steps_interval, 
      data = intervalav_steps, 
      type = 'l', 
      geom=c("line"),
      xlab = "Interval", 
      ylab = "Number of steps", 
      main = "") +
  facet_wrap(~ type_of_day, ncol = 1)
```

CONCLUSION: There is a different pattern of steps for weekdays and weekends. It can be seen that the activity is more spread during the day during the weekends, are more concentrated during weekdays